\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-11-08 k-planar.tex
Almost done......

\begin{proof}
Let $G$ be a $1$-plane multigraph. We may assume that no two edges incident to a common vertex of $G$ cross, since such a crossing can be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}. We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If two vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no two such edges  appear on the boundary of a common face. 
\note{DW}{Add ``So each face is bounded by a triangle, and this process terminates by Euler's formula.''}

\note{DW}{I think $G_0$ should be replaced by $G'$ throughout this proof.}

@@

Refer to \cref{one_planar_example}.
To understand the structure of $G$, it is helpful to consider the plane graph $G'$  obtained by replacing each pair of edges $vw$ and $xy$ that cross at $p$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \defin{dummy vertex} $p$. Let $F$ be a face of $G_0$ and let $W:=w_0,\ldots,w_{r-1}$ be the facial walk around $F$. In the following paragraphs, subscripts on the vertices in $W$ are implicitly taken modulo $r$. By edge-maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$.

\note{DW}{Put \cref{one_planar_example} here?}

Consider some pair of edges $vw$ and $xy$ that cross at point $p$, which is a degree-4 vertex of $G'$.  Since no pair of edges incident to a common vertex cross each other, $v$, $w$, $x$, and $y$ are all distinct.  There are four faces $F_1,\ldots,F_4$ of $G'$ with $p$ on their boundary and, from the preceding discussion each of $vx$, $xw$, $wy$, and $yv$ is an uncrossed edge of $G$.  Therefore, by removing each pair of crossing edges from $G$ we obtain a plane multigraph $G_0$ each of whose faces is bounded by three or four edges.  Therefore $G$ is a $4$-framed graph with frame $G_0$.
\end{proof}

@@

If $G_0$ has a face $f$ of length 2, then add a new vertex inside $f$ adjacent to both vertices on the boundary of $f$, which creates two new triangular faces $f_1$ and $f_2$. If $f$ is a lake, then make $f_1$ and $f_2$ lakes. If $F$ is a nation, then make $f_1$ a nation and make $f_2$ a lake. The resulting map graph is still $G$. So we may assume that $G_0$ is an edge-maximal multigraph with no faces of length 2 (and with each face labelled a nation or a lake), such that $G$ is the corresponding map graph. This is well-defined since the assumption of having no face of length 2 implies that $|E(G_0|\leq 3|V(G)|-6$.

Suppose that some face $f$ of $G_0$ has a disconnected boundary (implying $G_0$ is disconnected). Let $v$ and $w$ be vertices in distinct components of the boundary of $f$. Add the edge $vw$ to $G_0$ across $f$. The corresponding map graph is unchanged, which contradicts the edge-maximality of $G_0$. Thus no face of $G_0$ is disconnected. Suppose that some face $f$ of $G_0$ has a repeated vertex in the boundary walk of $f$. Let $u,v,w$ be consecutive vertices on the boundary of $f$. So $u,v,w$ are distinct and $uw$ is not an edge in the boundary of $f$. Add the edge $uw$ across $f$ and consider the resulting face $uvw$ to be a lake. Since $v$ appears elsewhere in the boundary of $f$, the corresponding map graph is unchanged, which contradicts the edge-maximality of $G_0$. Thus no facial walk of $G_0$ has a repeated vertex. Since each facial walk is connected, every face of $G_0$ is bounded by a cycle, and $G_0$ is biconnected.

>>>>

\begin{proof}
Let $G$ be a (connected) $1$-plane multigraph. We may assume that no two edges incident to a common vertex of $G$ cross, since such a crossing can be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}. We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If distinct vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no newly-introduced edge $vw$ is on the boundary of a face with an edge parallel to $vw$.  Since $G$ is connected, each edge added during this process splits a face $F$ with a facial walk of length at least $4$ into two faces each having a shorter facial walk than that of $F$.
Therefore this process eventually reaches a stage where each face is either bounded by a $3$-cycle or (a pre-existing) $2$-cycle and the process ends.

% \note{DW}{I think $G_0$ should be replaced by $G'$ throughout this proof.}
% \note{PM}{Agreed. Done.}

To understand the structure of $G$, it is helpful to consider the plane graph $G'$  obtained by replacing each pair of edges $vw$ and $xy$ that cross at $p$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \defin{dummy vertex} $p$. Let $F$ be a face of $G'$ and let $W:=w_0,\ldots,w_{r-1}$ be the facial walk around $F$. In the following paragraphs, subscripts on the vertices in $W$ are implicitly taken modulo $r$. By edge-maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$.

@@

Refer to \cref{one_planar_example}.  Consider some pair of edges $vw$ and $xy$ that cross at point $p$, which is a degree-4 vertex of $G'$.  Since no pair of edges incident to a common vertex cross each other, $v$, $w$, $x$, and $y$ are all distinct.  There are four faces $F_1,\ldots,F_4$ of $G'$ with $p$ on their boundary and, from the preceding discussion each of $vx$, $xw$, $wy$, and $yv$ is an uncrossed edge of $G$.  Therefore, by removing each pair of crossing edges from $G$ we obtain a plane multigraph $G_0$ each of whose faces is bounded by three or four edges.  Therefore $G$ is a $4$-framed graph with frame $G_0$.
\end{proof}

@@

\note{PM}{Here we're using $f$ to denote a face, earlier we were using $F$.}
If $G_0$ has a face $f$ of length 2, then add a new vertex inside $f$ adjacent to both vertices on the boundary of $f$, which creates two new triangular faces $f_1$ and $f_2$. If $f$ is a lake, then make $f_1$ and $f_2$ lakes. If $F$ is a nation, then make $f_1$ a nation and make $f_2$ a lake. The resulting map graph is still $G$. So we may assume that $G_0$ is an edge-maximal multigraph with no faces of length 2 (and with each face labelled a nation or a lake), such that $G$ is the corresponding map graph. This is well-defined since the assumption of having no face of length 2 implies that $|E(G_0|\leq 3|V(G_0)|-6$.

Suppose that some face $f$ of $G_0$ has a disconnected boundary (implying that $G_0$ is disconnected). Let $v$ and $w$ be vertices in distinct components of the boundary of $f$. Add the edge $vw$ to $G_0$ across $f$. The corresponding map graph is unchanged, which contradicts the edge-maximality of $G_0$. Thus no face of $G_0$ is disconnected.

Suppose that some face $f$ of $G_0$ has a repeated vertex $v$ in the boundary walk of $f$ and let $u,v,w$ be consecutive vertices in this boundary walk.  Since $v$ appears more than once in the boundary walk of $f$, the degree of $v$ is greater than $1$.  Since the degree of $v$ is greater than $1$ and the length of $f$ is greater than $2$, $u\neq w$. Since $v$ is repeated, $uw$ is not an edge in the boundary of $f$. Add the edge $uw$ across $f$ and consider the resulting face $uvw$ to be a lake. Since $v$ appears elsewhere in the boundary of $f$, the corresponding map graph is unchanged, which contradicts the edge-maximality of $G_0$. Thus no facial walk of $G_0$ has a repeated vertex. Since each facial walk is connected, each face of $G_0$ is bounded by a cycle, and $G_0$ is biconnected.

### 2022-11-08 response2.tex
Almost done......

2. Are the nations incident to a vertex counted with multiplicity in dmap graphs? The definition does not say so, but the proof of Lemma 27 (as currently stated) assumes not only this, but that the embedded graph under consideration is 2-connected. It also seems that frames are loopless, so slightly more care is called for when they are constructed.
It may be worth noting explicitly that you allow d-framed graphs to
have edges outside the outer face of their frame - I was unsure about
this given the phrase “interior of the face”.

\tba

@@

16. page 20, first paragraph of Lemma 25. Well, parallel edges can appear
on the boundary of a common face if they were there before. Add a
sentence about why the process ends.

\textcolor{orange}{DW: I have suggested a sentence to add. }

\tba

@@

\textcolor{orange}{DW: I think $G_0$ should be replaced by $G'$ throughout this proof.}

\tba

>>>>

2. Are the nations incident to a vertex counted with multiplicity in dmap graphs? The definition does not say so, but the proof of Lemma 27 (as currently stated) assumes not only this, but that the embedded graph under consideration is 2-connected.  It also seems that frames are loopless, so slightly more care is called for when they are constructed.

\begin{response}
  We have addressed this by adding the assumption that the underlying planar graph $G_0$ is edge-maximal. More precisely, it is not possible to add an edge to $G_0$ (and relabel faces as nation or lake) to obtain the same map graph $G$.  We then argue that this implies that $G_0$ is biconnected. 
\end{response}

It may be worth noting explicitly that you allow d-framed graphs to
have edges outside the outer face of their frame - I was unsure about
this given the phrase “interior of the face”.

@@

16. page 20, first paragraph of Lemma 25. Well, parallel edges can appear
on the boundary of a common face if they were there before.

\begin{response}
  We rephrased this to clarify that it only applies to newly-introduced edges.
\end{response}

Add a sentence about why the process ends.

@@

\begin{response}
  We have moved Figure 4 a bit later and, indeed, several occurrences of $G_0$ should have been $G'$.  These are corrected now.
\end{response}
